DNA Methylation Hierarchical Variance Model\nPaper Figures

### 2021-01-05 Paper_Figures.Rmd
Removed extraneous code, minor code edits

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdir))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))

# # plot beta/M values
# all_beta = FullAnnotation[,-c(1,2,3,4,5,6,7,8,9)]
# mean_beta = rowMeans(all_beta)
# all_M = logit(all_beta)
# mean_M = rowMeans(all_M)
# p_beta <- data.frame(value=mean_beta) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.01,fill="lightgray",color="grey50") + theme_classic() + xlim(c(0,1)) + theme(aspect.ratio = 1) + xlab("Beta value")
# p_M <- data.frame(value=mean_M) %>% gather %>% ggplot() + geom_histogram(aes(x=value),binwidth=0.1,fill="lightgray",color="grey50") + theme_classic() + xlim(c(-5,5)) + theme(aspect.ratio = 1) + xlab("M value")
# p_BM <- ggarrange(p_beta, p_M, nrow = 1, ncol = 2)
# # ggsave("betaMdists.svg",p_BM)
```

@@

\subsection{Prior Probabilities}
Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution: gamma(2,2).
\newline Additionally: Prior relaxation was performed by algebraically increasing the variance of the distributions (or each Gaussian component in the case of mu) while maintaining the same mode (peak).

@@

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

@@

\newpage
\subsection{Figure 2A: Example site fit, 2B: Posteriors}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/example_posteriors.png)
\newpage
\section{Figure 2: Model Fit Results}
\subsection{Parameter Distributions}
Histograms are median values of posterior distribution
```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))

@@

# plot mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))

@@

\subsection{Conservation score}
```{r Fig1B, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}

post_histogram <- function(postvals,mask,bw,xl,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..count../length(mask),fill=gene_assoc), binwidth=bw) +
    ggtitle(titlestr) + xlim(xl) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
}

# overlaid transparent histograms
post_histogram <- function(postvals,mask,bw,titlestr) {
  # stacked histogram colored by gene-associated sites
  # and prior curve overlaid
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  gene_assoc_df <- plot_df %>% filter(gene_assoc)
  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
  p1
}

# plot the two conservations score prob sums
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median,genemask,0.05,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-3,3),breaks=c(-3,-2,-1,0,1,2,3))
print(p_scores)
# ggsave("CpGscores_overlay.svg")

# plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median)) + geom_histogram(aes(y=..count..), binwidth=0.05, fill="lightgray",color = "grey50") + xlim(c(-1,2.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
# ggsave("gene_scores_bootstrap.svg")
```

@@

@@

\subsection{Regulatory regions}
Table comparing different groups of CpG sites:
```{r Table1, echo=FALSE, warning=FALSE}

#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5

prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(c(1,4,2,5,3,6)) %>% # for island, N shore, S shore, N shelf, S shelf, sea
  rename(`CpG Conservation Score`=score_median) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, `CpG Conservation Score`), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variances", columns = vars(sigmaP, sigmaPT, sigmaT)) %>% cols_align("center")

# stats for cpg island relations
# summarize mean of each group, do kruskal wallis and post hoc Tukey
my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
  group_by(ir = Relation_to_UCSC_CpG_Island) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir))
my_data %>%
  summarise_at(vars(score_median), mean, na.rm = TRUE)
kruskal.test(score_median ~ ir, data = my_data)
#install.packages("MultNonParam")
library(MultNonParam)
tukey.kruskal.test(my_data$score_median, my_data$ir, alpha = 0.001)

# stats for functional regions
summarize_conservation <- function(data,str) {
  indsin <- str_detect(data$UCSC_RefGene_Group, str) & !is.na(data$UCSC_RefGene_Group)
  indsout <- !str_detect(data$UCSC_RefGene_Group, str) | is.na(data$UCSC_RefGene_Group)
  if (str=="any") {
    indsin <- !is.na(data$UCSC_RefGene_Group)
    indsout <- is.na(data$UCSC_RefGene_Group)
  }
  data.frame(cons_in=mean(data$score_median[indsin], na.rm=TRUE),
         cons_out=mean(data$score_median[indsout],na.rm=TRUE),
         nsites=sum(indsin),
         pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
}

bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,pval)) %>%
  gt() %>%
  fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)

# new stats
FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
  mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
  mutate(FR=if_else(is.na(FR),"NA",FR)) %>%
  mutate(iTSS200 = str_detect(FR,"TSS200"),
         i5UTR = str_detect(FR,"5'UTR|5UTR|5URT"),
         iBody = str_detect(FR,"Body"),
         i3UTR = str_detect(FR,"3'UTR|3UTR"),
         iTSS1500 = str_detect(FR,"TSS1500"),
         i1stEx = str_detect(FR,"1stExon"),
         iExBnd = str_detect(FR,"ExonBnd"))
fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
#fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR, data = FRdata)
summary(fr.lm)

library(rcompanion)
fr.srh = scheirerRayHare(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
```

\newpage
\subsection{Figure 1: CpG Distance Analysis}
Corrected for gene direction, based on Ensembl gene database \newline
```{r Fig1, echo=FALSE, warning=FALSE, message=FALSE, fig.width=7, fig.height=3, comment=NA}
# load ensembl db
library(EnsDb.Hsapiens.v75) # ensembl annotation with gene position (hg19 version)
edb <- EnsDb.Hsapiens.v75
geneKeys <- keys(edb, keytype="GENENAME") %>% intersect(unique(gened_data$UCSC_RefGene_Name))
edbData<- AnnotationDbi::select(edb, keys=geneKeys, keytype="GENENAME",
                                columns=c("GENEBIOTYPE", "GENESEQSTART", "GENESEQEND", "SEQNAME", "SEQSTRAND")) %>%
  filter(SEQNAME %in% c(as.character(1:22),"X","Y")) %>%
  mutate(GENESTART = GENESEQSTART*(SEQSTRAND==1) + GENESEQEND*(SEQSTRAND==-1)) %>%
  select(-c(GENESEQSTART,GENESEQEND))

# calculate CpG distance based on Ensembl gene start position and direction
distanced_data <- gened_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  filter(UCSC_RefGene_Name %in% geneKeys) %>%
  mutate(EDBCHR=edbData$SEQNAME[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBSTART=edbData$GENESTART[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBDIR=edbData$SEQSTRAND[match(UCSC_RefGene_Name,edbData$GENENAME)],
         EDBTYPE=edbData$GENEBIOTYPE[match(UCSC_RefGene_Name,edbData$GENENAME)]
         ) %>%
  filter(CHR==EDBCHR) %>%
  mutate(cpgdist = if_else(EDBDIR==1, MAPINFO-EDBSTART, EDBSTART-MAPINFO))

distance_length <- distanced_data %>%
  filter(!is.na(UCSC_RefGene_Name)) %>%
  select(CHR, UCSC_RefGene_Name, cpgdist, c(sigmaP:sigmaT,score_median:score_reg)) %>%
  mutate(location_bin = cut(cpgdist, c(seq(-1500, 20000, by = 100)), include.lowest = TRUE)) %>%
  group_by(location_bin) %>%
  summarise(mean_sigmaP = mean(sigmaP), mean_sigmaPT = mean(sigmaPT), mean_sigmaT = mean(sigmaT),  mean_score_median = mean(score_median), mean_score_median2 = mean(score_median2), n_obs = n(), .groups='drop') %>%
  mutate(location_bin = as.numeric(str_extract(location_bin, "(?<=,)(.*)(?=])")))

distance_length %>%
  drop_na() %>%
  pivot_longer(mean_sigmaP:mean_sigmaT) %>% 
  mutate(name = factor(name, c("mean_sigmaP","mean_sigmaPT","mean_sigmaT"), c("SigmaP","SigmaPT","SigmaT"))) %>%
  sample_frac() %>%
  ggplot(aes(location_bin, value, color = name, fill = name)) +
  geom_point(key_glyph = draw_key_rect) +
  ggforce::facet_zoom(location_bin < 5000) +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = NULL,
       fill = NULL,
       color = NULL,
       x = "Distance from gene start site (binned to 100 bp)",
       y = "Average Sigma Value", 
       caption = "Figure 1: Average Sigma values of single CpGs as a function of position relative to gene.") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
cat("\n")

# again for conservation scores
distance_length %>%
  drop_na() %>%
  pivot_longer(mean_score_median) %>% 
  mutate(name = factor(name, c("mean_score_median"), c("Score_median"))) %>%
  sample_frac() %>%
  ggplot(aes(location_bin, value, color = name, fill = name)) +
  geom_point(key_glyph = draw_key_rect) +
  ggforce::facet_zoom(location_bin < 5000) +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = NULL,
       fill = NULL,
       color = NULL,
       x = "Distance from gene start site (binned to 100 bp)",
       y = "Average Conservation Score", 
       caption = "Figure 1: Average conservation scores of single CpGs as a function of position relative to gene.") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
cat("\n")

# and look at number of CpG in each bin
distance_length %>%
  drop_na() %>%
  pivot_longer(n_obs) %>%
  mutate(name = factor(name, c("n_obs"), c("n_obs"))) %>%
  sample_frac() %>%
  ggplot(aes(location_bin, value, color = name, fill = name)) +
  geom_point(key_glyph = draw_key_rect) +
  ggforce::facet_zoom(location_bin < 5000) +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(title = NULL,
       fill = NULL,
       color = NULL,
       x = "Distance from gene start site (binned to 100 bp)",
       y = "Number of CpGs", 
       caption = "Figure 1: Frequency of single CpGs as a function of position relative to gene start site.") +
  scale_color_brewer(palette = "Set2") +
  scale_fill_brewer(palette = "Set2")
```

>>>>

prepdatafolder = "data_final_rg3"
cat(sprintf("Load prepped data dir: %s (needs to be pre-processed)...",prepdatafolder))
gened_data <- read_csv(here(prepdatafolder, "gened_data.csv"), col_types = list("CHR" = col_character()))
prepped_data <- read_csv(here(prepdatafolder, "prepped_data.csv"))
permed_data <- read_csv(here(prepdatafolder, "permed_data.csv"))

@@

\subsection{Prior Probabilities (Fig S1)}
Priors for each parameter (mu, nu, sigmaP/PT/T) were defined using TCGA 450k array data from colorectal tumor patients with paired multiple tumor and/or normal+tumor samples. \newline Empirical values of mean and standard deviations were assesed for all 450k CpG sites. \newline The resulting distributions were fit with: bimodal gaussian (mu), Cauchy (betaT), or gamma distributions (sigmas). \newline
SigmaE (error variance) prior was defined using a non-informative gamma distribution: gamma(2,2).
\newline Additionally: Prior relaxation was performed by algebraically increasing the variance of the distributions (or each Gaussian component in the case of mu) while maintaining the same mode (peak).

@@

# plot TCGA estimates for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with prior fits overlaid
pM <- post_prior_histogram(mu_ests, function(x) fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), function(x) fBG(x,0.3557,-3.0675, sqrt(3)*0.75214,1.1290, sqrt(3)*1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_ests, function(x) dcauchy(x,-0.04051, 0.19330), function(x) dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "Nu")
pP <- post_prior_histogram(sigmaP_ests, function(x) dgamma(x, 5.1989, 16.6192), function(x) dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_ests, function(x) dgamma(x, 3.5539, 5.5324), function(x) dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_ests, function(x) dgamma(x, 6.8383, 25.7560), function(x) dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT",legFlag=TRUE)
pE <- ggplot() + stat_function(fun=function(x) dgamma(x, 2, 2),geom="line",color="blue") + ggtitle("SigmaE") + xlim(c(0,2)) + theme_classic() + theme(legend.position = "none") + xlab("SigmaE") + ylab("Freq.")

@@

\newpage
\subsection{Figure 2A: Example site fit, 2B: Posteriors}
![Fit](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_fit.png)
\newline \newline
![Posteriors](/Users/kevinmurgas/Documents/Data+ project/Manuscript/Paper Figures (screenshots)/example_posteriors.png)

\newpage
\section{Model Fit Results}
\subsection{Fig 3C: Parameter Distributions}
Histograms are median values of posterior distribution
```{r Params_site, echo=FALSE, fig.fullwidth=TRUE, warning=FALSE, comment=NA}
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
cat(sprintf("N sites: total=%i, gene-associated=%i, non-gene=%i  \n",length(genemask),sum(genemask),length(genemask)-sum(genemask)))

@@

# plot hierarchical model posterior medians for mu,betaT,sigmaE,sigmaP,sigmaPT,sigmaT; with priors overlaid
pM <- post_prior_histogram(mu_full$p50, genemask, function(x) 1.5*fBG(x,0.3557,-3.0675, 0.75214,1.1290, 1.39197), 0.1, c(-5,5), "Mu")
pB <- post_prior_histogram(betaT_full$p50, genemask, function(x) 1.7*dcauchy(x,-0.04051, 0.5799), 0.1, c(-5,5), "BetaT")
pE <- post_prior_histogram(sigmaE_full$p50, genemask, function(x) 4*dgamma(x,2,2), 0.05, c(0,2), "SigmaE")
pP <- post_prior_histogram(sigmaP_full$p50, genemask, function(x) 2*dgamma(x, 2.7693, 7.0029), 0.05, c(0,2), "SigmaP")
pPT <- post_prior_histogram(sigmaPT_full$p50, genemask, function(x) 1.8*dgamma(x, 2.1457, 2.4819), 0.05, c(0,2), "SigmaPT")
pT <- post_prior_histogram(sigmaT_full$p50, genemask, function(x) 1.2*dgamma(x, 3.3643, 10.430), 0.05, c(0,2), "SigmaT")
# log sigmas
pLP <- post_prior_histogram(log(sigmaP_full$p50), genemask, function(x) 0.52*dgamma(exp(x), 2.7693, 7.0029), 0.1, c(-5,5), "log(SigmaP)")
pLPT <- post_prior_histogram(log(sigmaPT_full$p50), genemask, function(x) 0.7*dgamma(exp(x), 2.1457, 2.4819), 0.1, c(-5,5), "log(SigmaPT)")
pLT <- post_prior_histogram(log(sigmaT_full$p50), genemask, function(x) 0.4*dgamma(exp(x), 3.3643, 10.430), 0.1, c(-5,5), "log(SigmaT)") + theme(legend.position = c(0.8, 0.6), legend.text=element_text(size=6), legend.title=element_text(size=6))

@@

@@

\subsection{Fig 3A: Conservation score}
```{r Fig3A, echo=FALSE, fig.height=2, fig.width=6.5, fig.fullwidth=TRUE, warning=FALSE}

# staggered (dodge) histograms
post_histogram <- function(postvals,mask,bw,titlestr) {
  # stacked histogram colored by gene-associated sites
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(aes(y=..density..,fill=gene_assoc), binwidth=bw, position="dodge") +
    ggtitle(titlestr) + theme_classic() + theme(legend.position = "none") + xlab(titlestr) + ylab("Freq.")
}

# overlaid transparent histograms
post_histogram <- function(postvals,mask,bw,titlestr) {
  # transparent histogram colored by gene-associated sites
  plot_df <- data.frame(val=postvals, gene_assoc=mask)
  gene_assoc_df <- plot_df %>% filter(gene_assoc)
  not_gene_assoc_df <- plot_df %>% filter(!gene_assoc)
  p1 <- ggplot(plot_df, aes(x=val)) +
    geom_histogram(data=not_gene_assoc_df, aes(y=..density..,fill="#00BFC4",alpha=0.4), binwidth=bw) +
    geom_histogram(data=gene_assoc_df, aes(y=..density..,fill="#F8766D",alpha=0.4), binwidth=bw) +
    ggtitle(titlestr) + xlab(titlestr) + ylab("Density") + theme_classic() 
  p1
}

# plot the two conservations score prob sums (using /log(2) to convert log_e to log_2)
genemask = !(FullAnnotation$UCSC_RefGene_Name=="")
p_scores <- post_histogram(CpGscore_full$median/log(2),genemask,0.05,"CpG Conservation score") +
  theme(aspect.ratio=1/4,legend.position = "none") +
  scale_x_continuous(limits=c(-4,4),breaks=c(-4,-2,0,2,4)) + 
  scale_y_continuous(breaks=c(0,0.1,0.2,0.3,0.4,0.5))
print(p_scores)
# ggsave("CpGscores_overlay.svg")

# Fig 4B: plot gene scores and adjusted bootstrap distribution
p1 <- permed_data %>% ggplot(aes(x=score_median/log(2))) + geom_histogram(aes(y=..count..), binwidth=0.1, fill="lightgray",color = "grey50") + xlim(c(-1.5,3.5)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p2 <- permed_data %>% ggplot(aes(x=score_median_v3)) + geom_histogram(aes(y=..count..), binwidth=0.02, fill="lightgray", color="grey50") + xlim(c(0,1)) + theme_classic() + ylab("Number of genes") + theme(aspect.ratio = 1)
p_genes <- ggarrange(p1,p2,nrows=2,ncol=1)
# ggsave("gene_scores_bootstrap.svg")
```

@@

\subsection{CpG Island Relation (Fig 3B)}
Table comparing different groups of CpG sites based on CpG Island relationship: Island, Shore, Shelf, Sea
```{r CpGIslandRelation, echo=FALSE, warning=FALSE}
#, include=TRUE, results='asis', fig.keep='all', fig.show='asis', fig.height=3, fig.width=5

# for a table summary of sigmas and conservation score
prepped_data %>%
  group_by(`Island relation` = Relation_to_UCSC_CpG_Island) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  summarise_at(vars(sigmaP:sigmaT,score_median), mean, na.rm = TRUE) %>%
  dplyr::rename(ir = `Island relation`) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", "South "),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", "North "),
                        TRUE ~ ir)) %>%
  dplyr::rename(`Island relation` = ir) %>%
  arrange(c(1,4,2,5,3,6)) %>% # for island, N shore, S shore, N shelf, S shelf, sea
  rename(`CpG Conservation Score`=score_median) %>%
  gt() %>%
  fmt_number(columns = vars(sigmaP, sigmaPT, sigmaT, `CpG Conservation Score`), decimals = 3) %>%
  tab_spanner(label = "Hierarchical Variances", columns = vars(sigmaP, sigmaPT, sigmaT)) %>% cols_align("center")

# stats for cpg island relations
my_data = prepped_data %>% select(score_median, Relation_to_UCSC_CpG_Island) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  group_by(ir = Relation_to_UCSC_CpG_Island) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir))
# summarize mean of each group and convert to fold change
my_data %>% summarise_at(vars(score_median), mean, na.rm = TRUE) %>% mutate(fc = 2^score_median)
# run Kruskal Wallis ANOVA and post-hoc Tukey
kruskal.test(score_median ~ ir, data = my_data)
#install.packages("MultNonParam")
library(MultNonParam)
tukey.kruskal.test(my_data$score_median, my_data$ir, alpha = 0.001)

# Fig 3B: plot CpG island relation score distributions
prepped_data %>% select(Name, Relation_to_UCSC_CpG_Island, score_median) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  dplyr::rename(ir = Relation_to_UCSC_CpG_Island, score=score_median) %>%
  mutate(ir = case_when(is.na(ir) ~ "Sea",
                        str_detect(ir, "^S") ~ str_replace(ir, "^S_", ""),
                        str_detect(ir, "^N") ~ str_replace(ir, "^N_", ""),
                        TRUE ~ ir)) %>%
  mutate(ir = factor(ir, levels = c("Island", "Shore", "Shelf", "Sea"))) %>%
  #mutate(ir = factor(ir, levels = c("Sea", "Shelf", "Shore", "Island"))) %>%
  ggplot() + geom_violin(aes(x=ir,y=score,fill=ir),draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4)) + theme(aspect.ratio=1/4,legend.position = "none")
# ggsave("CpGscore_islands_violin.svg")
```

\subsection{Functional Regions (Fig 3C)}
Table comparing different groups of CpG sites based on functional regions: 5'UTR, TSS1500, TSS200, 1stExon, Body, ExonBnd, 3'UTR
```{r FunctionalRegions, echo=FALSE, warning=FALSE}
# stats for functional regions
summarize_conservation <- function(data,str) {
  indsin <- str_detect(data$UCSC_RefGene_Group, str) & !is.na(data$UCSC_RefGene_Group)
  indsout <- !str_detect(data$UCSC_RefGene_Group, str) | is.na(data$UCSC_RefGene_Group)
  if (str=="any") {
    indsin <- !is.na(data$UCSC_RefGene_Group)
    indsout <- is.na(data$UCSC_RefGene_Group)
  }
  data.frame(cons_in=mean(data$score_median[indsin], na.rm=TRUE),
         cons_out=mean(data$score_median[indsout],na.rm=TRUE),
         nsites=sum(indsin),
         nout=sum(indsout),
         pval=wilcox.test(data$score_median[indsin],data$score_median[indsout])$p.value)
}

temp <- bind_rows({prepped_data %>% summarize_conservation("any") %>% mutate(region="Gene-Associated")},
          {prepped_data %>% summarize_conservation("TSS1500") %>% mutate(region="TSS1500")},
          {prepped_data %>% summarize_conservation("TSS200") %>% mutate(region="TSS200")},
          {prepped_data %>% summarize_conservation("5'UTR|5UTR|5URT") %>% mutate(region="5'UTR")},
          {prepped_data %>% summarize_conservation("1stExon") %>% mutate(region="1stExon")},
          {prepped_data %>% summarize_conservation("ExonBnd") %>% mutate(region="ExonBnd")},
          {prepped_data %>% summarize_conservation("Body") %>% mutate(region="Body")},
          {prepped_data %>% summarize_conservation("3'UTR|3UTR") %>% mutate(region="3'UTR")}) %>%
  select(c(region,cons_in,cons_out,nsites,nout,pval)) %>%
  mutate(cons_in = cons_in/log(2), cons_out = cons_out/log(2)) %>% # change of base e to base 2
  mutate(fold_change = 2^cons_in, fold_diff = 2^(cons_in - cons_out)) # calculating fold difference (2^ (log2 difference))
temp %>% gt() %>% fmt_number(columns = vars(cons_in,cons_out,pval), decimals = 3)

# multiple regression stats (using log2 score)
FRdata = data.frame(CpGscore = log2(prepped_data$sigmaP / prepped_data$sigmaT) ) %>%
  mutate(FR = pull(prepped_data,UCSC_RefGene_Group)) %>%
  mutate(FR=if_else(is.na(FR),"NA",FR)) %>%
  mutate(iTSS200 = str_detect(FR,"TSS200"),
         i5UTR = str_detect(FR,"5'UTR|5UTR|5URT"),
         iBody = str_detect(FR,"Body"),
         i3UTR = str_detect(FR,"3'UTR|3UTR"),
         iTSS1500 = str_detect(FR,"TSS1500"),
         i1stEx = str_detect(FR,"1stExon"),
         iExBnd = str_detect(FR,"ExonBnd"))

# lm model
fr.lm = lm(formula = CpGscore ~ iTSS200 + i5UTR + iBody + i3UTR + iTSS1500 + i1stEx + iExBnd, data = FRdata)
summary(fr.lm)

# Fig 3C: plot functional region score distributions
temp <- prepped_data %>% select(Name, UCSC_RefGene_Group, score_median) %>%
  mutate(score_median = score_median / log(2) ) %>% # convert log_e to log_2
  separate_rows(UCSC_RefGene_Group, sep=";") %>%
  distinct() %>%
  dplyr::rename(fr = UCSC_RefGene_Group, score=score_median) %>%
  mutate(fr = case_when(fr=="" ~ "NA",
                        str_detect(fr, "5URT") ~ "5'UTR",
                        str_detect(fr, "3UTR") ~ "3'UTR",
                        TRUE ~ fr)) %>%
  distinct()
#keepregions = c("TSS1500","Body", "3'UTR","1stExon","TSS200","5'UTR","ExonBnd")
keepregions = c("TSS200","1stExon", "5'UTR","TSS1500","Body","3'UTR","ExonBnd")
temp %>%
  mutate(fr = factor(fr, levels = keepregions)) %>%
  filter(fr %in% keepregions) %>%
  ggplot() + geom_violin(aes(x=fr,y=score,fill=fr), draw_quantiles = 0.5) + theme_classic() + scale_y_continuous(limits=c(-5,5),breaks=c(-4,-2,0,2,4,6)) + theme(aspect.ratio=1/4,legend.position = "none") + geom_hline(yintercept = 0, color="black", alpha=0.2)
# ggsave("CpGscore_regions_violin.svg")
#rm(temp)
